---
output: html_document
editor_options: 
  chunk_output_type: console
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup}
knitr::opts_chunk$set(echo = FALSE)

# Load your packages
library(tidyverse) 
library(here) 
library(cowplot)
library(lubridate) 
library(zoo)
library(trend)
library(sf)
library(mapview); mapviewOptions(fgb = FALSE) 
library(cowplot)
library(agricolae)
library(knitr)

# Set your working directory
here()

# Set your ggplot theme
mytheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right")
theme_set(mytheme)

# Load your datasets
WQ_data_raw <- read.csv(here("./Raw_Data/WQ_data.txt"), header = TRUE, sep = "|",
                        colClasses = c("ProgramLocationID"="factor",
                                       "Year"="factor",
                                       "Month"="factor",
                                       "ParameterName"="factor"))

```


# Rationale and Research Questions
Urban coastal ecosystems are sensitive to stormwater runoff, which is the rain water and waste water that flows over streets, sidewalks, and other places. It picks up debris, fertilizers, and other pollutants along the way before entering into the stormwater system. The system is designed to drain cities of rain water and help to minimize flooding. Therefore, the stormwater systems carries the things that water picks up along the way before depositing the runoff to the surrounding waterways. This creates pollution problems and other serious impacts to the living species in the ecosystem, which happened near the Biscayne Bay area. To address such issues, the city of Miami Beach implemented the Miami Beach fertilizer ordinance that prohibits the use of fertilizers containing nitrogen and phosphorus during the summer rainy season. The ordinance was passed in January, 2021, and was adopted in April of the same year. The ordinance is effective from May 15  through November 1 every year (accounting for raining season). 

This study examines whether measurable changes in nutrient concentrations or related indicators appeared following the implementation of the Miami Beach fertilizer ordinance by using data from the City of Miami Beach Water Monitoring Program. The goal is to quantify these changes through time series and statistical analyses. We chose to look at this specific dataset because it includes measurements for multiple years, and it’s specific to the area/location that we are looking to assess. It also includes data points for various nutrients measurements that are crucial to our analyses, such as values for total nitrogen, total phosphorus, dissolved oxygen, and PH.   

Question 1: How did water quality in Biscayne Bay change following the 2021 summer fertilizer ordinance in terms of A) Total Phosphorous, B) Total Nitrogen, C) Dissolved Oxygen, and D) pH?

Question 2: 
 


\newpage

# Dataset Information

```{r echo=FALSE}
#wrangling

#Insert underscores into parameter names
levels(WQ_data_raw$ParameterName) <- c("Dissolved_O2", "NO2_NO3_Filtered", "pH", 
                                    "Salinity", "Specific_Conductivity", 
                                    "Total_Kjeldahl_Nitrogen", "Total_Nitrogen",
                                    "Total_Phosphorous", "Turbidity", 
                                    "Water_Temperature")

#pivot parameteres
WQ_parameters_raw <- WQ_data_raw %>%
  #select  columns of interest
  select(SampleDate, ParameterName, ResultValue, Year, Month, ProgramLocationID,
         OriginalLatitude, OriginalLongitude) %>%
  group_by(SampleDate, Year, Month, ProgramLocationID, OriginalLatitude,
           OriginalLongitude, ParameterName) %>%
  summarise(
    mean_ResultValue = mean(ResultValue)) %>%
  #move parameters of interest into their own columns
  pivot_wider(names_from = ParameterName, values_from = mean_ResultValue) %>%
  mutate(SampleDate = ymd_hms(SampleDate),
         Group = ifelse(SampleDate < as.Date("2021-04-1"), "Pre", "Post"))


#monthly averages data frame (for time series)
WQ_monthly_data <- WQ_parameters_raw %>%
  group_by(Year, Month, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-",Month)))


#Find missing observations (needed for interpolation)
dates_by_month <- as.data.frame(seq.Date(from=as.Date("2016/08/01"),
                                      to=as.Date("2024/12/01"),
                                      by="month"))
colnames(dates_by_month) <- "Date"
WQ_monthly_complete <- left_join(dates_by_month, WQ_monthly_data)


#interpolate monthly data
WQ_monthly_complete <- WQ_monthly_complete %>%
  mutate(
    Interp_Phosphorous = na.approx(Mean_Phosphorous),
    Interp_pH = na.approx(Mean_pH),
    Interp_Dissolved_O2 = na.approx(Mean_Dissolved_O2),
    Interp_Total_N = na.approx(Mean_Total_N),
    Group = ifelse(Date < as.Date("2021-04-01"), "Pre", "Post"))


#yearly numbers
yearly_averages <- WQ_parameters_raw %>%
  group_by(Year, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))

yearly_averages_bysite <- WQ_parameters_raw %>%
  group_by(Year, ProgramLocationID, OriginalLatitude, OriginalLongitude) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))


#split data into pre- and post-ordinance data frames
pre_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Pre")
  
post_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Post")

pre_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Pre")

post_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Post")


#parameter summary table
Minima <- c(
  min(WQ_parameters_raw$Total_Phosphorous, na.rm=TRUE),
  min(WQ_parameters_raw$Total_Nitrogen, na.rm=TRUE),
  min(WQ_parameters_raw$Dissolved_O2, na.rm=TRUE),
  min(WQ_parameters_raw$pH, na.rm=TRUE))

Maxima <- c(
  max(WQ_parameters_raw$Total_Phosphorous, na.rm=TRUE),
  max(WQ_parameters_raw$Total_Nitrogen, na.rm=TRUE),
  max(WQ_parameters_raw$Dissolved_O2, na.rm=TRUE),
  max(WQ_parameters_raw$pH, na.rm=TRUE))

Means <- c(
  mean(WQ_parameters_raw$Total_Phosphorous, na.rm=TRUE),
  mean(WQ_parameters_raw$Total_Nitrogen, na.rm=TRUE),
  mean(WQ_parameters_raw$Dissolved_O2, na.rm=TRUE),
  mean(WQ_parameters_raw$pH, na.rm=TRUE))

SDs <- c(
  sd(WQ_parameters_raw$Total_Phosphorous, na.rm=TRUE),
  sd(WQ_parameters_raw$Total_Nitrogen, na.rm=TRUE),
  sd(WQ_parameters_raw$Dissolved_O2, na.rm=TRUE),
  sd(WQ_parameters_raw$pH, na.rm=TRUE))

parameter.summary <- data.frame(
  "Parameter" = c("Total Phosphorous", "Total Nitrogen", "Dissolved Oxygen", "pH"),
  "Units" = c("mg/L", "mg/L", "mg/L", "NA"),
  "Minimum" = Minima,
  "Maximum" = Maxima,
  "Mean" = Means,
  "Std_Deviation" = SDs)

kable(
  parameter.summary,
  align = c("l","c","r","r","r","r"),
  digits = c(1,1,5,4,2,2),
  caption = "Parameter Summary Statistics")
```

\newpage

# Exploratory Analysis 

```{r echo=FALSE}
#Creating simple features data frame for spatial visualization 
SampleLocations_sf <- WQ_data_raw %>%
  distinct(ProgramLocationID,OriginalLongitude, OriginalLatitude, .keep_all = FALSE) %>%
  st_as_sf(
    coords = c("OriginalLongitude", "OriginalLatitude"),
    crs= 4326)

#Visualization of sample collection locations 
mapviewOptions(basemaps = "Esri.WorldImagery")

mapview(
  SampleLocations_sf,
  layer.name= "Sample Locations",
  legend= FALSE,
  col.regions="blue",
  col="yellow",
  alpa.regions= 0.5
  )
```

```{r echo=FALSE}
#Exploring series

phosphorous.preview <- ggplot(WQ_monthly_complete) +
  geom_line(aes(x=Date, y=Mean_Phosphorous)) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Dissolved Phosphorous (mg/L)",
    x="")

nitrogen.preview <- ggplot(WQ_monthly_complete) +
  geom_line(aes(x=Date, y=Mean_Total_N)) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Total Nitrogen (mg/L)",
    x="")

oxygen.preview <- ggplot(WQ_monthly_complete) +
  geom_line(aes(x=Date, y=Mean_Dissolved_O2)) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="")

pH.preview <- ggplot(WQ_monthly_complete) +
  geom_line(aes(x=Date, y=Mean_pH)) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="pH",
    x="Date")

plot_grid(phosphorous.preview, nitrogen.preview, oxygen.preview, pH.preview, nrow=4)

```
Exploratory box plots showing the distribution of parameters of interest by year
individually and as a combined plot 
```{r}

Total_Nitrogen_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Interp_Total_N)
            )+
  geom_boxplot()+
  xlab("Year")+
  ylab("Total Nitrogen (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())
print(Total_Nitrogen_Distribution)

Total_Phosphorous_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Interp_Phosphorous)
            )+
  geom_boxplot()+
  xlab("Year")+
  ylab("TotalPhosphorous (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())
print(Total_Phosphorous_Distribution)

Dissolved_O2_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Interp_Dissolved_O2)
            )+
  geom_boxplot()+
  xlab("Year")+
  ylab("Dissolved O2 (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())
print(Dissolved_O2_Distribution)

pH_distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Interp_pH)
            )+
  geom_boxplot()+
  xlab("Year")+
  ylab("pH")

print(pH_distribution)

CombinedDistributionPlot <- plot_grid(
  Total_Nitrogen_Distribution,
  Total_Phosphorous_Distribution,
  Dissolved_O2_Distribution,
  pH_distribution,
  ncol=1,
  align = "v"
)

print(CombinedDistributionPlot)
```


\newpage

# Analysis

```{r}
#time series

#monthly

monthly.P.ts <- ts(WQ_monthly_complete$Interp_Phosphorous, start=c(2016,8), 
                   frequency=12)
monthly.P.decomp <- stl(monthly.P.ts, s.window="periodic")
plot(monthly.P.decomp)
monthly.P.components <- as.data.frame(monthly.P.decomp$time.series[,1:3])
monthly.P.components <- mutate(monthly.P.components,
                               observed = WQ_monthly_complete$Interp_Phosphorous,
                               date = WQ_monthly_complete$Date)

monthly.N.ts <- ts(WQ_monthly_complete$Interp_Total_N, start=c(2016,8), 
                   frequency=12)
monthly.N.decomp <- stl(monthly.N.ts, s.window="periodic")
plot(monthly.N.decomp)
monthly.N.components <- as.data.frame(monthly.N.decomp$time.series[,1:3])
monthly.N.components <- mutate(monthly.N.components,
                               observed = WQ_monthly_complete$Interp_Total_N,
                               date = WQ_monthly_complete$Date)

monthly.O.ts <- ts(WQ_monthly_complete$Interp_Dissolved_O2, start=c(2016,8),
                   frequency=12)
monthly.O.decomp <- stl(monthly.O.ts, s.window="periodic")
plot(monthly.O.decomp)
monthly.O.components <- as.data.frame(monthly.O.decomp$time.series[,1:3])
monthly.O.components <- mutate(monthly.O.components,
                               observed = WQ_monthly_complete$Interp_Dissolved_O2,
                               date = WQ_monthly_complete$Date)

monthly.pH.ts <- ts(WQ_monthly_complete$Interp_pH, start=c(2016,8),
                   frequency=12)
monthly.pH.decomp <- stl(monthly.pH.ts, s.window="periodic")
plot(monthly.pH.decomp)
monthly.pH.components <- as.data.frame(monthly.pH.decomp$time.series[,1:3])
monthly.pH.components <- mutate(monthly.pH.components,
                               observed = WQ_monthly_complete$Interp_pH,
                               date = WQ_monthly_complete$Date)

#yearly
yearly.P.ts <- ts(yearly_averages$Mean_Phosphorous, 
                  start = 2016,
                  frequency = 1)
#yearly.P.decomp <- stl(yearly.P.ts, s.window = "periodic") (not working)
plot(yearly.P.ts)

yearly.N.ts <- ts(yearly_averages$Mean_Total_N, 
                  start = 2016,
                  frequency = 1)
plot(yearly.N.ts)

yearly.O.ts <- ts(yearly_averages$Mean_Dissolved_O2, 
                  start = 2016,
                  frequency = 1)
plot(yearly.O.ts)

yearly.pH.ts <- ts(yearly_averages$Mean_pH, 
                  start = 2016,
                  frequency = 1)
plot(yearly.pH.ts)

```


```{r}
#ANOVA
yearly.P.anova <- aov(data=WQ_monthly_complete, Mean_Phosphorous ~ Year)
summary(yearly.P.anova)
TukeyHSD(yearly.P.anova)
P.HSD <- HSD.test(yearly.P.anova, "Year", group=TRUE)
P.HSD

yearly.N.anova <- aov(data=WQ_monthly_complete, Mean_Total_N ~ Year)
summary(yearly.N.anova)
TukeyHSD(yearly.N.anova)
N.HSD <- HSD.test(yearly.N.anova, "Year", group=TRUE)
N.HSD

yearly.O.anova <- aov(data=WQ_monthly_complete, Mean_Dissolved_O2 ~ Year)
summary(yearly.O.anova)
TukeyHSD(yearly.O.anova)
O.HSD <- HSD.test(yearly.O.anova, "Year", group=TRUE)
O.HSD

yearly.pH.anova <- aov(data=WQ_monthly_complete, Mean_pH ~ Year)
summary(yearly.pH.anova)
TukeyHSD(yearly.pH.anova)
O.HSD <- HSD.test(yearly.pH.anova, "Year", group=TRUE)
O.HSD

```

## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
https://data.florida-seacar.org/programs/details/4058 
