---
title: "Trend Analysis of Water Quality Data in Biscayne Bay: Impacts of Summer Fertilizer Ordinance"
author: "Dashiell Ridolfi-Starr & Joy Wu & Jordan Heinlein"
date: "Fall 2025"
 
output:
  bookdown::html_document2:
    toc: TRUE
    fig_caption: yes
    lof: true
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---
\newpage
\tableofcontents 
\newpage

## List of Tables
*Dataset Information*

1. Table \@ref(tab:table1): Parameter Summary Statistics
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load packages
library(tidyverse) 
library(here) 
library(cowplot)
library(lubridate) 
library(zoo)
library(trend)
library(sf)
library(mapview); mapviewOptions(fgb = FALSE) 
library(cowplot)
library(agricolae)
library(knitr)

# Set working directory
#here()
# Set ggplot theme
mytheme <- theme_classic(base_size = 11) +
  theme(axis.text = element_text(color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right")
theme_set(mytheme)

# Load datasets
WQ_data_raw <- read.csv(here("./Raw_Data/WQ_data.txt"), header = TRUE, sep = "|",
                        colClasses = c("ProgramLocationID"="factor",
                                       "Year"="factor",
                                       "Month"="factor",
                                       "ParameterName"="factor"))

```


# Rationale and Research Questions
Urban coastal ecosystems are sensitive to stormwater runoff, which is rain water and waste water that flows over streets, sidewalks, and other surfaces. It picks up debris, oil, fertilizers, and other pollutants along the way before entering the stormwater system. The system is designed to drain cities of rain water and help minimize flooding. As a result, carries and deposits polluted runoff into surrounding waterways, causing harm to nearby ecossytems. Unfortunately, this is the case for Biscayne Bay near the City of Miami Beach. To address such issues, Miami-Dade County implemented an ordinance that prohibits the use of fertilizers containing nitrogen and phosphorus during the summer rainy season. The ordinance was passed in January, 2021, and was adopted in April of the same year. The ordinance is effective from May 15 through October 31 every year (accounting for raining season). 

This study examines whether measurable changes in nutrient concentrations or related indicators appeared following the implementation of the Miami-Dade County fertilizer ordinance by using data from the City of Miami Beach Water Monitoring Program. The goal is to quantify these changes through time series and statistical analyses. We chose to look at this specific dataset because it includes measurements for multiple years, and it is specific to the location that we are looking to assess. It also measures parameters that are crucial to our analyses, such as total nitrogen, total phosphorus, dissolved oxygen, and pH. 


\newpage

# Dataset Information

```{r wrangling, echo=FALSE, message=FALSE, warning=FALSE}
#wrangling

#Insert underscores into parameter names
levels(WQ_data_raw$ParameterName) <- c("Dissolved_O2", "NO2_NO3_Filtered", "pH", 
                                    "Salinity", "Specific_Conductivity", 
                                    "Total_Kjeldahl_Nitrogen", "Total_Nitrogen",
                                    "Total_Phosphorous", "Turbidity", 
                                    "Water_Temperature")

#pivot parameteres
WQ_parameters_raw <- WQ_data_raw %>%
  #select  columns of interest
  select(SampleDate, ParameterName, ResultValue, Year, Month, ProgramLocationID,
         OriginalLatitude, OriginalLongitude) %>%
  group_by(SampleDate, Year, Month, ProgramLocationID, OriginalLatitude,
           OriginalLongitude, ParameterName) %>%
  summarise(
    mean_ResultValue = mean(ResultValue)) %>%
  #move parameters of interest into their own columns
  pivot_wider(names_from = ParameterName, values_from = mean_ResultValue) %>%
  mutate(SampleDate = ymd_hms(SampleDate),
         Group = ifelse(SampleDate < as.Date("2021-04-1"), "Pre", "Post"))

#remove outliers from dissolved O2 and pH
WQ_parameters_treated <- WQ_parameters_raw  %>%
  mutate(
    pH = ifelse(pH>15,NA,pH),
    Dissolved_O2 = ifelse(Dissolved_O2>700,NA,Dissolved_O2))

#monthly averages data frame (for time series)
WQ_monthly_data <- WQ_parameters_treated %>%
  group_by(Year, Month, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate_all(~ifelse(is.nan(.),NA,.)) %>%
  mutate(Date = ym(paste0(Year,"-",Month)))
  

#Find missing observations (needed for interpolation)
dates_by_month <- as.data.frame(seq.Date(from=as.Date("2016/08/01"),
                                      to=as.Date("2024/12/01"),
                                      by="month"))
colnames(dates_by_month) <- "Date"
WQ_monthly_complete <- left_join(dates_by_month, WQ_monthly_data)


#interpolate monthly data
WQ_monthly_complete <- WQ_monthly_complete %>%
  select(
    Date, Month, Mean_Phosphorous, Mean_Total_N, Mean_Dissolved_O2, Mean_pH) %>%
  mutate(
    Year = as.factor(year(Date)),
    Interp_Phosphorous = na.approx(Mean_Phosphorous),
    Interp_pH = na.approx(Mean_pH),
    Interp_Dissolved_O2 = na.approx(Mean_Dissolved_O2),
    Interp_Total_N = na.approx(Mean_Total_N),
    Group = ifelse(Date < as.Date("2021-04-01"), "Pre", "Post"))


#yearly numbers
yearly_averages <- WQ_parameters_treated %>%
  group_by(Year, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))

yearly_averages_bysite <- WQ_parameters_treated %>%
  group_by(Year, ProgramLocationID, OriginalLatitude, OriginalLongitude) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))


#split data into pre- and post-ordinance data frames
pre_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Pre")
  
post_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Post")

pre_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Pre")

post_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Post")


#parameter summary table
Minima <- c(
  min(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  min(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  min(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  min(WQ_parameters_treated$pH, na.rm=TRUE))

Maxima <- c(
  max(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  max(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  max(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  max(WQ_parameters_treated$pH, na.rm=TRUE))

Means <- c(
  mean(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  mean(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  mean(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  mean(WQ_parameters_treated$pH, na.rm=TRUE))

SDs <- c(
  sd(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  sd(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  sd(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  sd(WQ_parameters_treated$pH, na.rm=TRUE))

parameter.summary <- data.frame(
  "Parameter" = c("Total Phosphorous", "Total Nitrogen", "Dissolved Oxygen", "pH"),
  "Units" = c("mg/L", "mg/L", "mg/L", "NA"),
  "Minimum" = Minima,
  "Maximum" = Maxima,
  "Mean" = Means,
  "Std_Deviation" = SDs)

P.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Total_Phosphorous))+
  labs(
    x= "Date",
    y="Total Phosphorous (mg/L)")
N.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Total_Nitrogen))+
  labs(
    x="Date",
    y="Total Nitrogen (mg/L)")
O.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Dissolved_O2))+
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="Date")
pH.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=pH))+
  labs(
    y="pH",
    x="Date")

points.preview <- plot_grid(P.points, N.points, O.points, pH.points, nrow=2, ncol=2)
```

[Insert dataset description]

Below are the structure of the raw dataset used in this project and initial plots of parameter values.
```{r data structure, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Water Quality Parameter Readings"}
#displaying data

#raw data structure
str(WQ_data_raw)
```
\
\
```{r figure1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Water Quality Parameter Readings"}
#show points 
print(points.preview)
```

## Data Wrangling

[insert wrangling description]

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
#paramter summary
kable(
  parameter.summary,
  align = c("l","c","r","r","r","r"),
  digits = c(1,1,4,4,3,3),
  caption = "Table 1. Parameter Summary Statistics")
```


\newpage

# Exploratory Analysis 

## Geographic Context

```{r map, echo=FALSE, message=FALSE, warning=FALSE}
#Creating simple features data frame for spatial visualization 
SampleLocations_sf <- WQ_data_raw %>%
  distinct(ProgramLocationID,OriginalLongitude, OriginalLatitude, .keep_all = FALSE) %>%
  st_as_sf(
    coords = c("OriginalLongitude", "OriginalLatitude"),
    crs= 4326)

#Visualization of sample collection locations 
mapviewOptions(basemaps = "Esri.WorldImagery")

mapview(
  SampleLocations_sf,
  layer.name= "Sample Locations",
  legend= FALSE,
  col.regions="blue",
  col="yellow",
  alpa.regions= 0.5
  )
```

[insert description of study area, anything else relevant]
\
## Monthly Averages

Monthly averages of the study parameters are visualized below. The vertical dotted line represents the implementation of the Miami-Dade County summer fertilizer ordinance (April, 2021).
```{r preview plots, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.cap="Monthly Averages of Water Quality Parameters"}
#Exploring series

phosphorous.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Phosphorous)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Total Phosphorous (mg/L)",
    x="")

nitrogen.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Total_N)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Total Nitrogen (mg/L)",
    x="")

oxygen.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Dissolved_O2)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="")

pH.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_pH)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="pH",
    x="Date")

plot_grid(phosphorous.preview, nitrogen.preview, oxygen.preview, pH.preview, nrow=4, align='v')

```

[insert description of above plots]
\
## Yearly Averages

Exploratory box plots showing the distribution of parameters of interest by year individually and as a combined plot 
```{r boxplots, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.cap="Distributions of Water Quality Parameter Values by Year"}

Total_Nitrogen_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=Year,y=Mean_Total_N)
            )+
  geom_boxplot(na.rm= TRUE)+
  xlab("Year")+
  ylab("Total Nitrogen (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())

Total_Phosphorous_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_Phosphorous)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("TotalPhosphorous (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())


Dissolved_O2_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_Dissolved_O2)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("Dissolved O2 (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())


pH_distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_pH)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("pH")


CombinedDistributionPlot <- plot_grid(
  Total_Nitrogen_Distribution,
  Total_Phosphorous_Distribution,
  Dissolved_O2_Distribution,
  pH_distribution,
  ncol=1,
  align = "v"
)

print(CombinedDistributionPlot)
```

[insert description of boxplots]
\
\newpage

# Analysis
## Question 1: How do the pre- and post-ordinance trends in monthly averages compare in terms of A) Total Phosphorous, B) Total Nitrogen, C) Dissolved Oxygen, and D) pH?


###	Pre- vs. Post-Ordinance Monthly Trend Analysis


```{r pre vs post TS, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.cap="Comparing Pre- and Post-Ordinance Trends in Water Quality Parameters"}
#time series

#P overall monthly ts
monthly.P.ts <- ts(WQ_monthly_complete$Interp_Phosphorous, start=c(2016,8), 
                   frequency=12)
monthly.P.decomp <- stl(monthly.P.ts, s.window="periodic")
monthly.P.components <- as.data.frame(monthly.P.decomp$time.series[,1:3])
monthly.P.components <- mutate(monthly.P.components,
                               observed = WQ_monthly_complete$Interp_Phosphorous,
                               date = WQ_monthly_complete$Date)

#N overall monthly ts
monthly.N.ts <- ts(WQ_monthly_complete$Interp_Total_N, start=c(2016,8), 
                   frequency=12)
monthly.N.decomp <- stl(monthly.N.ts, s.window="periodic")
monthly.N.components <- as.data.frame(monthly.N.decomp$time.series[,1:3])
monthly.N.components <- mutate(monthly.N.components,
                               observed = WQ_monthly_complete$Interp_Total_N,
                               date = WQ_monthly_complete$Date)


#O2 overall monthly ts
monthly.O.ts <- ts(WQ_monthly_complete$Interp_Dissolved_O2, start=c(2016,8),
                   frequency=12)
monthly.O.decomp <- stl(monthly.O.ts, s.window="periodic")
monthly.O.components <- as.data.frame(monthly.O.decomp$time.series[,1:3])
monthly.O.components <- mutate(monthly.O.components,
                               observed = WQ_monthly_complete$Interp_Dissolved_O2,
                               date = WQ_monthly_complete$Date)

#pH overall monthly ts
monthly.pH.ts <- ts(WQ_monthly_complete$Interp_pH, start=c(2016,8),
                   frequency=12)
monthly.pH.decomp <- stl(monthly.pH.ts, s.window="periodic")
monthly.pH.components <- as.data.frame(monthly.pH.decomp$time.series[,1:3])
monthly.pH.components <- mutate(monthly.pH.components,
                               observed = WQ_monthly_complete$Interp_pH,
                               date = WQ_monthly_complete$Date)



#P pre vs. post monthly trend analysis
P.pre.ts <- ts(pre_ordinance_monthly$Interp_Phosphorous, start = c(2016,8),
               frequency=12)
P.pre.decomp <- stl(P.pre.ts, s.window="periodic") 
P.pre.components <- as.data.frame(P.pre.decomp$time.series[,1:3])
P.pre.components <- mutate(P.pre.components,
                            observed = pre_ordinance_monthly$Interp_Phosphorous,
                            date = pre_ordinance_monthly$Date)
P.pre.smk <- smk.test(P.pre.ts) #no significant trend


P.post.ts <- ts(post_ordinance_monthly$Interp_Phosphorous, start = c(2021,4),
                frequency=12)
P.post.decomp <- stl(P.post.ts, s.window = "periodic")
P.post.components <- as.data.frame(P.post.decomp$time.series[,1:3])
P.post.components <- mutate(P.post.components,
                            observed = post_ordinance_monthly$Interp_Phosphorous,
                            date = post_ordinance_monthly$Date)
P.post.smk <- smk.test(P.post.ts) #significant trend! 

#P pre vs post plot
P.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Phosphorous)) +
  geom_line(data=P.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2018-10-01"), y=0.02,
            label=paste0("SMK p-value = ", round(P.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=P.post.components, aes(x=date, y=observed)) +
  geom_line(data=P.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=0.04,
            label=paste0("SMK p-value = ", round(P.post.smk$p.value, 3), "*"),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Total Phosphorous (mg/L)",
    x="")

#N pre vs. post trend analysis
N.pre.ts <- ts(pre_ordinance_monthly$Interp_Total_N, start = c(2016,8),
               frequency=12)
N.pre.decomp <- stl(N.pre.ts, s.window="periodic") 
N.pre.components <- as.data.frame(N.pre.decomp$time.series[,1:3])
N.pre.components <- mutate(N.pre.components,
                            observed = pre_ordinance_monthly$Interp_Total_N,
                            date = pre_ordinance_monthly$Date)
N.pre.smk <- smk.test(N.pre.ts) 


N.post.ts <- ts(post_ordinance_monthly$Interp_Total_N, start = c(2021,4),
                frequency=12)
N.post.decomp <- stl(N.post.ts, s.window = "periodic")
N.post.components <- as.data.frame(N.post.decomp$time.series[,1:3])
N.post.components <- mutate(N.post.components,
                            observed = post_ordinance_monthly$Interp_Total_N,
                            date = post_ordinance_monthly$Date)
N.post.smk <- smk.test(N.post.ts)

#N pre vs post plot
N.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Total_N)) +
  geom_line(data=N.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2018-10-01"), y=0.8,
            label=paste0("SMK p-value = ", round(N.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=N.post.components, aes(x=date, y=observed)) +
  geom_line(data=N.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-01-01"), y=0.8,
            label=paste0("SMK p-value = ", round(N.post.smk$p.value, 3)),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Total Nitrogen (mg/L)",
    x="")

#O2 pre vs. post trend analysis
O.pre.ts <- ts(pre_ordinance_monthly$Interp_Dissolved_O2, start = c(2016,8),
               frequency=12)
O.pre.decomp <- stl(O.pre.ts, s.window="periodic") 
O.pre.components <- as.data.frame(O.pre.decomp$time.series[,1:3])
O.pre.components <- mutate(O.pre.components,
                            observed = pre_ordinance_monthly$Interp_Dissolved_O2,
                            date = pre_ordinance_monthly$Date)
O.pre.smk <- smk.test(O.pre.ts) 


O.post.ts <- ts(post_ordinance_monthly$Interp_Dissolved_O2, start = c(2021,4),
                frequency=12)
O.post.decomp <- stl(O.post.ts, s.window = "periodic")
O.post.components <- as.data.frame(O.post.decomp$time.series[,1:3])
O.post.components <- mutate(O.post.components,
                            observed = post_ordinance_monthly$Interp_Dissolved_O2,
                            date = post_ordinance_monthly$Date)
O.post.smk <- smk.test(O.post.ts)

#O2 pre vs post plot
O.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Dissolved_O2)) +
  geom_line(data=O.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2019-09-01"), y=45,
            label=paste0("SMK p-value = ",round(O.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=O.post.components, aes(x=date, y=observed)) +
  geom_line(data=O.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=45,
            label=paste0("SMK p-value = ", round(O.post.smk$p.value, 3)),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="")

#pH pre vs. post trend analysis
pH.pre.ts <- ts(pre_ordinance_monthly$Interp_pH, start = c(2016,8),
               frequency=12)
pH.pre.decomp <- stl(pH.pre.ts, s.window="periodic") 
pH.pre.components <- as.data.frame(pH.pre.decomp$time.series[,1:3])
pH.pre.components <- mutate(pH.pre.components,
                            observed = pre_ordinance_monthly$Interp_pH,
                            date = pre_ordinance_monthly$Date)
pH.pre.smk <- smk.test(pH.pre.ts) 


pH.post.ts <- ts(post_ordinance_monthly$Interp_pH, start = c(2021,4),
                frequency=12)
pH.post.decomp <- stl(pH.post.ts, s.window = "periodic")
pH.post.components <- as.data.frame(pH.post.decomp$time.series[,1:3])
pH.post.components <- mutate(pH.post.components,
                            observed = post_ordinance_monthly$Interp_pH,
                            date = post_ordinance_monthly$Date)
pH.post.smk <- smk.test(pH.post.ts)

#pH pre vs post plot
pH.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_pH)) +
  geom_line(data=pH.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2019-03-01"), y=8.75,
            label=paste0("SMK p-value = ",round(pH.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=pH.post.components, aes(x=date, y=observed)) +
  geom_line(data=pH.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=9.5,
            label=paste0("SMK p-value = ", round(pH.post.smk$p.value, 3),"*"),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  ylim(6.5,10) +
  labs(
    y="pH")

#combined pre vs post trend analysis plot
plot_grid(P.prevspost.plot, N.prevspost.plot, O.prevspost.plot, pH.prevspost.plot,
          nrow=4)
```
\
```{r SMK, echo=FALSE, message=FALSE, warning=FALSE}
#pre vs post SMK results summary table
pre.stat.summary <- c(P.pre.smk$statistic, N.pre.smk$statistic, 
                   O.pre.smk$statistic, pH.pre.smk$statistic)
pre.pvalue.summary <- c(P.pre.smk$p.value, N.pre.smk$p.value,
                        O.pre.smk$p.value, pH.pre.smk$p.value)
post.stat.summary <- c(P.post.smk$statistic, N.post.smk$statistic,
                       O.post.smk$statistic, pH.post.smk$statistic)
post.pvalue.summary <- c(P.post.smk$p.value, N.post.smk$p.value,
                        O.post.smk$p.value, pH.post.smk$p.value)

prevspost.SMK.summary <- data.frame(
  "Pre-Ordinance_Z.Stat" = pre.stat.summary,
  "Pre-Ordinance_p-value" = pre.pvalue.summary,
  "Post-Ordinance_Z.Stat" = post.stat.summary,
  "Post-Ordinance_p-value" = post.pvalue.summary,
  row.names = c("Total_Phosphorous", "Total_Nitrogen", "Dissolved_Oxygen", "pH"))

#smk summary table
kable(prevspost.SMK.summary, 
      digits=c(3,3,3,3),
      caption="Results Summary of Pre- and Post-Ordinance Monthly Averages")
```

We conducted monthly time series analysis of the various parameters to assess whether monotonic trends differed before and after the implementation of the ordinance. The dashed vertical line in these plots represent the ordinance implementation time. Seasonal Mann-Kendall (SMK) tests were applied separately to pre- and post-ordinance periods to evaluate the presence of statistically significant monotonic trends while accounting for seasonality (Figure xx).   

Time series analysis reveals statistically significant, downward monotonic trends in post-ordinance Total Phosphorous (SMK; z < 0, p-value < 0.05) and pH (SMK; z < 0, p-value < 0.05). For both parameters, no significant trend was observed prior to the ordinance implementation. For the other two parameters, dissolved oxygen and pH, plotted time series trends show some variability, but no pre- or post-ordinance variables showed significant monotonic trends. SMK results for all parameters are summarized in Table __.


###	Comparison of Monthly and Yearly Trends

To assess whether the interpolated monthly observation reflect accurate description of the data, we created plots that show trend of the yearly mean value for the different parameters between 2016 and 2025 (Figure xx).  For all parameters, yearly averages maintained the direction and relative magnitude of post-ordinance changes observed earlier in the monthly data.
```{r yearly vs monthly TS, echo=FALSE, message=FALSE, warning=FALSE}
#yearly TS with combined plots
par(mfrow = c(2,2))
yearly.P.ts <- ts(yearly_averages$Mean_Phosphorous, 
                  start = 2016,
                  frequency = 1)
plot(yearly.P.ts,
     xlab = "Year",
     ylab = "Phosphorous (mg/L)")

yearly.N.ts <- ts(yearly_averages$Mean_Total_N, 
                  start = 2016,
                  frequency = 1)
plot(yearly.N.ts,
     xlab = "Year",
     ylab = "Nitrogen (mg/L)")

yearly.O.ts <- ts(yearly_averages$Mean_Dissolved_O2, 
                  start = 2016,
                  frequency = 1)
plot(yearly.O.ts,
      xlab = "Year",
     ylab = "Dissolved Oxygen (mg/L)")

yearly.pH.ts <- ts(yearly_averages$Mean_pH, 
                  start = 2016,
                  frequency = 1)
plot(yearly.pH.ts,
     xlab = "Year",
     ylab = "pH")
par(mfrow = c(1,1))

#comparison of monthly interpolated TS & yearly non-interpolated TS
#Phosphorous
P.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Phosphorous)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Phosphorous (mg/L)",
    title = "Yearly Mean Total Phosphorous")
  
plot_grid(P.prevspost.plot, P.yearly.plot,
          nrow=2,
          align = "v")

#Nitrogen
N.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Total_N)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Nitrogen (mg/L)",
    title = "Yearly Mean Total Nitrogen")


plot_grid(N.prevspost.plot, N.yearly.plot,
          nrow=2,
          align = "v")

O.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Dissolved_O2)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Dissolved Oxygen (mg/L)",
    title = "Yearly Mean Total Dissolved Oxygen")


plot_grid(O.prevspost.plot, O.yearly.plot,
          nrow=2,
          align = "v")

pH.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_pH)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "pH",
    title = "Yearly Mean pH")


plot_grid(pH.prevspost.plot, pH.yearly.plot,
          nrow=2,
          align = "v")
```

To compare the results in a more visual way, we stacked the monthly and yearly plots to assess their trends for each of the parameters. For all four parameters, the monthly and yearly plots seem to follow similar trends. The yearly mean total phosphorus concentrations show a clear decline beginning in 2021, which is consistent with the significant post-ordinance monotonic trend we identified earlier in monthly analyses. The yearly mean total nitrogen presented larger variability regarding trends throughout the years, but its general trend had a declining direction after 2021. The yearly mean dissolved oxygen had large spikes during the early times and showed limited differentiation after ordinance implementation, while the yearly mean pH showed a modest decline following 2021.

The similar plot patterns between the monthly and yearly time series suggest that the observed post-ordinance declines in phosphorous and pH are consistent across time scales, while nitrogen and dissolved oxygen show high variability in both datasets. 



## Question 2: Are there significant differences among the yearly averages for A) Total Phosphorous, B) Total Nitrogen, C) Dissolved Oxygen, and D) pH? 

```{r anova, echo=FALSE, message=FALSE, warning=FALSE}
#ANOVA
yearly.P.anova <- aov(data=WQ_monthly_complete, Mean_Phosphorous ~ Year)
#TukeyHSD(yearly.P.anova)
P.HSD <- HSD.test(yearly.P.anova, "Year", group=TRUE)

yearly.N.anova <- aov(data=WQ_monthly_complete, Mean_Total_N ~ Year)
#TukeyHSD(yearly.N.anova)
N.HSD <- HSD.test(yearly.N.anova, "Year", group=TRUE)

yearly.O.anova <- aov(data=WQ_monthly_complete, Mean_Dissolved_O2 ~ Year)
#TukeyHSD(yearly.O.anova)
O.HSD <- HSD.test(yearly.O.anova, "Year", group=TRUE)

yearly.pH.anova <- aov(data=WQ_monthly_complete, Mean_pH ~ Year)
#TukeyHSD(yearly.pH.anova)
pH.HSD <- HSD.test(yearly.pH.anova, "Year", group=TRUE)

#anova results table
P.anova.results <- summary(yearly.P.anova)[[1]]
N.anova.results <- summary(yearly.N.anova)[[1]]
O.anova.results <- summary(yearly.O.anova)[[1]]
pH.anova.results <- summary(yearly.pH.anova)[[1]]

anova.results <- bind_rows(
  P.anova.results,
  N.anova.results,
  O.anova.results,
  pH.anova.results)

anova.results <- anova.results %>%
  mutate(Parameter = c(
    "Total_Phosphorous", "Total_Phosphorous", "Total_Nitrogen", "Total_Nitrogen",
    "Dissolved_O2", "Dissolved_O2", "pH", "pH")) %>%
  select(Parameter, Df, `Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)`)
 
kable(anova.results,
      row.names = T,
      digits = c(1,1,3,3,3,5),
      caption="Summary of ANOVA Results")

#HSD results into table
P.groups <- as.data.frame(P.HSD$groups)
N.groups <- as.data.frame(N.HSD$groups)
O.groups <- as.data.frame(O.HSD$groups)
pH.groups <- as.data.frame(pH.HSD$groups)

parameter.anova.groups <- data.frame(
  "Phosphorous" = P.groups,
  "Nitrogen" = N.groups,
  "Dissolved_O2"= O.groups,
  "pH" = pH.groups)

kable(parameter.anova.groups,
      col.names=c("Year","Phosphorous_Mean","P_Group","Nitrogen_Mean","N_Group",
                  "Diss_O2_Mean","O2_Group","pH_Mean","pH_Group"),
            align=c("r","c","r","c","r","c","r","c"),
      digits=c(3,1,3,1,3,1,3,1,3),
      caption="Results Summary of Tukey's HSD Test on Yearly Averages")
```

One-way analysis of variance (ANOVA) was performed on the four study parameters using year as a categorical variable to calculate means. Analysis of Total Phosphorous (df = 8, F = 36.18, p < 0.00001) and Total Nitrogen (df = 8, F = 3.554, p < 0.005) found statistically significant differences among the yearly means for both parameters. There were no statistically significant differences among yearly means for Dissolved Oxygen (df = 8, F = 1.04, p = 0.413) or pH (df = 8, F = 1.648, p = 0.124). 

These results indicate that there are relatively stronger shifts between years in nutrient concentrations, particularly for phosphorous. On the other hand, physical water quality parameters presented weaker and more variable year-to-year differentiation. 

Following ANOVA, Tukey’s honestly significant difference test (HSD) was used to identify years with diverging means across variables. The results are summarized in Table __. Yearly Total Phosphorous means were more stratified, with seven years belonging to single, distinct mean groups. One statistically unique mean group occurs only before 2021, a second occurs only during 2021, and a third occurs only afterward. Total Nitrogen also showed statistically distinct mean groups occurring before and after 2021 but had less separation overall. Both parameters showed less separation among mean groups in 2017 and 2018. For dissolved oxygen and pH, the results indicated substantial overlap among yearly mean groups, which reinforce the earlier ANOVA results that suggested limited year-to-year differences for these parameters. 




\newpage

# Summary and Conclusions
### Total Phosphorous

\
### Total Nitrogen

\
### Dissolved Oxygen

\
### pH

### Summary

\
\newpage

# References
<add references here if relevant, otherwise delete this section> 
https://data.florida-seacar.org/programs/details/4058 
