---
title: "Trend Analysis of Water Quality Data in Biscayne Bay: Impacts of Summer Fertilizer Ordinance"
subtitle: https://github.com/jzwu01/Wu_Heinlein_RidolfiStarr
author: "Dashiell Ridolfi-Starr & Joy Wu & Jordan Heinlein"
date: "Fall 2025"
 
output:
  bookdown::html_document2:
    toc: TRUE
    fig_caption: yes
    lof: true
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

```{=tex}
\newpage
\tableofcontents 
\newpage
```
## List of Tables

*Dataset Information*

* Table \@ref(tab:table1): Parameter Summary Statistics

*Analysis*

* Table \@ref(tab:SMK): Results Summary of SMK on Pre- and Post-Ordinance Monthly Averages
* Table \@ref(tab:anova): Results Summary of ANOVA on Yearly Averages
* Table \@ref(tab:Tukey): Results Summary of Tukey's HSD on Yearly Averages

## List of Figures

*Dataset Information*

* Figure \@ref(fig:points): Water Quality Parameter Readings

*Exploratory Analysis*

* Figure \@ref(fig:previewplots): Monthly Averages of Water Quality Parameters
* Figure \@ref(fig:boxplots): Distributions of Water Quality Parameter Values by Year

*Analysis*

* Figure \@ref(fig:prevspostTS): Comparing Pre- and Post-Ordinance Trends in Water Quality Parameters
* Figure \@ref(fig:yearlyTS): Parameter Yearly Averages
* Figure \@ref(fig:Pyearlyvsmonthly): Comparing Yearly Averages with Interpolated Monthly Averages: Phosphorous
* Figure \@ref(fig:Nyearlyvsmonthly): Comparing Yearly Averages with Interpolated Monthly Averages: Nitrogen
* Figure \@ref(fig:Oyearlyvsmonthly): Comparing Yearly Averages with Interpolated Monthly Averages: Dissolved Oxygen
* Figure \@ref(fig:pHyearlyvsmonthly): Comparing Yearly Averages with Interpolated Monthly Averages: pH

    ```{=tex}
    \listoftables 
    \newpage
    \listoffigures 
    \newpage
    ```

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load packages
library(tidyverse) 
library(here) 
library(cowplot)
library(lubridate) 
library(zoo)
library(trend)
library(sf)
library(mapview); mapviewOptions(fgb = FALSE) 
library(cowplot)
library(agricolae)
library(knitr)

# Set working directory
#here()
# Set ggplot theme
mytheme <- theme_classic(base_size = 11) +
  theme(axis.text = element_text(color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right")
theme_set(mytheme)

# Load datasets
WQ_data_raw <- read.csv(here("./Raw_Data/WQ_data.txt"), header = TRUE, sep = "|",
                        colClasses = c("ProgramLocationID"="factor",
                                       "Year"="factor",
                                       "Month"="factor",
                                       "ParameterName"="factor"))

```

# Rationale and Research Questions

---

\newpage

# Dataset Information

## Raw Data

```{r wrangling, echo=FALSE, message=FALSE, warning=FALSE}
#wrangling

#Insert underscores into parameter names
levels(WQ_data_raw$ParameterName) <- c("Dissolved_O2", "NO2_NO3_Filtered", "pH", 
                                    "Salinity", "Specific_Conductivity", 
                                    "Total_Kjeldahl_Nitrogen", "Total_Nitrogen",
                                    "Total_Phosphorous", "Turbidity", 
                                    "Water_Temperature")

#pivot parameteres
WQ_parameters_raw <- WQ_data_raw %>%
  #select  columns of interest
  select(SampleDate, ParameterName, ResultValue, Year, Month, ProgramLocationID,
         OriginalLatitude, OriginalLongitude) %>%
  group_by(SampleDate, Year, Month, ProgramLocationID, OriginalLatitude,
           OriginalLongitude, ParameterName) %>%
  summarise(
    mean_ResultValue = mean(ResultValue)) %>%
  #move parameters of interest into their own columns
  pivot_wider(names_from = ParameterName, values_from = mean_ResultValue) %>%
  mutate(SampleDate = ymd_hms(SampleDate),
         Group = ifelse(SampleDate < as.Date("2021-04-1"), "Pre", "Post"))

#remove outliers from dissolved O2 and pH
WQ_parameters_treated <- WQ_parameters_raw  %>%
  mutate(
    pH = ifelse(pH>15,NA,pH),
    Dissolved_O2 = ifelse(Dissolved_O2>700,NA,Dissolved_O2))

#monthly averages data frame (for time series)
WQ_monthly_data <- WQ_parameters_treated %>%
  group_by(Year, Month, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate_all(~ifelse(is.nan(.),NA,.)) %>%
  mutate(Date = ym(paste0(Year,"-",Month)))
  

#Find missing observations (needed for interpolation)
dates_by_month <- as.data.frame(seq.Date(from=as.Date("2016/08/01"),
                                      to=as.Date("2024/12/01"),
                                      by="month"))
colnames(dates_by_month) <- "Date"
WQ_monthly_complete <- left_join(dates_by_month, WQ_monthly_data)


#interpolate monthly data
WQ_monthly_complete <- WQ_monthly_complete %>%
  select(
    Date, Month, Mean_Phosphorous, Mean_Total_N, Mean_Dissolved_O2, Mean_pH) %>%
  mutate(
    Year = as.factor(year(Date)),
    Interp_Phosphorous = na.approx(Mean_Phosphorous),
    Interp_pH = na.approx(Mean_pH),
    Interp_Dissolved_O2 = na.approx(Mean_Dissolved_O2),
    Interp_Total_N = na.approx(Mean_Total_N),
    Group = ifelse(Date < as.Date("2021-04-01"), "Pre", "Post"))


#yearly numbers
yearly_averages <- WQ_parameters_treated %>%
  group_by(Year, Group) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))

yearly_averages_bysite <- WQ_parameters_treated %>%
  group_by(Year, ProgramLocationID, OriginalLatitude, OriginalLongitude) %>%
  summarise(
    Mean_Phosphorous = mean(Total_Phosphorous, na.rm=TRUE),
    Mean_pH = mean(pH, na.rm=TRUE),
    Mean_Dissolved_O2 = mean(Dissolved_O2, na.rm=TRUE),
    Mean_Total_N = mean(Total_Nitrogen, na.rm=TRUE)) %>%
  mutate(Date = ym(paste0(Year,"-","01")))


#split data into pre- and post-ordinance data frames
pre_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Pre")
  
post_ordinance_monthly <- WQ_monthly_complete %>%
  filter(Group == "Post")

pre_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Pre")

post_ordinance_yearly <- yearly_averages %>%
  filter(Group == "Post")


#parameter summary table
Minima <- c(
  min(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  min(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  min(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  min(WQ_parameters_treated$pH, na.rm=TRUE))

Maxima <- c(
  max(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  max(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  max(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  max(WQ_parameters_treated$pH, na.rm=TRUE))

Means <- c(
  mean(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  mean(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  mean(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  mean(WQ_parameters_treated$pH, na.rm=TRUE))

SDs <- c(
  sd(WQ_parameters_treated$Total_Phosphorous, na.rm=TRUE),
  sd(WQ_parameters_treated$Total_Nitrogen, na.rm=TRUE),
  sd(WQ_parameters_treated$Dissolved_O2, na.rm=TRUE),
  sd(WQ_parameters_treated$pH, na.rm=TRUE))

parameter.summary <- data.frame(
  "Parameter" = c("Total Phosphorous", "Total Nitrogen", "Dissolved Oxygen", "pH"),
  "Units" = c("mg/L", "mg/L", "mg/L", "NA"),
  "Minimum" = Minima,
  "Maximum" = Maxima,
  "Mean" = Means,
  "Std_Deviation" = SDs)

P.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Total_Phosphorous))+
  labs(
    x= "Date",
    y="Total Phosphorous (mg/L)")
N.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Total_Nitrogen))+
  labs(
    x="Date",
    y="Total Nitrogen (mg/L)")
O.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=Dissolved_O2))+
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="Date")
pH.points <- ggplot() +
  geom_point(data=WQ_parameters_raw, aes(x=SampleDate, y=pH))+
  labs(
    y="pH",
    x="Date")

points.preview <- plot_grid(P.points, N.points, O.points, pH.points, nrow=2, ncol=2)
```

```{r data structure, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Water Quality Parameter Readings"}
#displaying data

#raw data structure
str(WQ_data_raw)
```

## Data Wrangling

```{r points, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Water Quality Parameter Readings"}
#show points 
print(points.preview)
```

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
#parameter summary
kable(
  parameter.summary,
  align = c("l","c","r","r","r","r"),
  digits = c(1,1,4,4,3,3),
  caption = "Parameter Summary Statistics")
```

---

\newpage

# Exploratory Analysis

## Geographic Context

```{r map, echo=FALSE, message=FALSE, warning=FALSE, label="map", fig.cap="Locations of Water Quality Sampling Sites"}
#Creating simple features data frame for spatial visualization 
SampleLocations_sf <- WQ_data_raw %>%
  distinct(ProgramLocationID,OriginalLongitude, OriginalLatitude, .keep_all = FALSE) %>%
  st_as_sf(
    coords = c("OriginalLongitude", "OriginalLatitude"),
    crs= 4326)

#Visualization of sample collection locations 
mapviewOptions(basemaps = "Esri.WorldImagery")

mapview(
  SampleLocations_sf,
  layer.name= "Sample Locations",
  legend= FALSE,
  col.regions="blue",
  col="yellow",
  alpa.regions= 0.5
  )
```

## Monthly Averages

```{r previewplots, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.cap="Monthly Averages of Water Quality Parameters"}
#Exploring series

phosphorous.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Phosphorous)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Total Phosphorous (mg/L)",
    x="")

nitrogen.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Total_N)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Total Nitrogen (mg/L)",
    x="")

oxygen.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_Dissolved_O2)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="")

pH.preview <- ggplot(WQ_monthly_complete,
  aes(x=Date, y=Mean_pH)) +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)+
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  labs(
    y="pH",
    x="Date")

plot_grid(phosphorous.preview, nitrogen.preview, oxygen.preview, pH.preview, nrow=4, align='v')

```

## Yearly Distributions

```{r boxplots, echo=FALSE, message=FALSE, warning=FALSE, fig.height=9, fig.cap="Distributions of Water Quality Parameter Values by Year"}

Total_Nitrogen_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=Year,y=Mean_Total_N)
            )+
  geom_boxplot(na.rm= TRUE)+
  xlab("Year")+
  ylab("Total Nitrogen (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())

Total_Phosphorous_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_Phosphorous)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("TotalPhosphorous (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())


Dissolved_O2_Distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_Dissolved_O2)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("Dissolved O2 (mg/L)")+
  theme(axis.title.x = element_blank())+
  theme(axis.text.x = element_blank())+
  theme(axis.ticks.x = element_blank())


pH_distribution <- ggplot(WQ_monthly_complete,
            aes(x=as.factor(Year),y=Mean_pH)
            )+
  geom_boxplot(na.rm = TRUE)+
  xlab("Year")+
  ylab("pH")


CombinedDistributionPlot <- plot_grid(
  Total_Phosphorous_Distribution,
  Total_Nitrogen_Distribution,
  Dissolved_O2_Distribution,
  pH_distribution,
  ncol=1,
  align = "v"
)

print(CombinedDistributionPlot)
```

---

\newpage

# Analysis

## Question 1: How do the pre- and post-ordinance trends in monthly averages compare in terms of A) Total Phosphorous, B) Total Nitrogen, C) Dissolved Oxygen, and D) pH?

### Pre- vs. Post-Ordinance Monthly Trend Analysis

```{r prevspostTS, echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.cap="Comparing Pre- and Post-Ordinance Trends in Water Quality Parameters"}
#time series

#P overall monthly ts
monthly.P.ts <- ts(WQ_monthly_complete$Interp_Phosphorous, start=c(2016,8), 
                   frequency=12)
monthly.P.decomp <- stl(monthly.P.ts, s.window="periodic")
monthly.P.components <- as.data.frame(monthly.P.decomp$time.series[,1:3])
monthly.P.components <- mutate(monthly.P.components,
                               observed = WQ_monthly_complete$Interp_Phosphorous,
                               date = WQ_monthly_complete$Date)

#N overall monthly ts
monthly.N.ts <- ts(WQ_monthly_complete$Interp_Total_N, start=c(2016,8), 
                   frequency=12)
monthly.N.decomp <- stl(monthly.N.ts, s.window="periodic")
monthly.N.components <- as.data.frame(monthly.N.decomp$time.series[,1:3])
monthly.N.components <- mutate(monthly.N.components,
                               observed = WQ_monthly_complete$Interp_Total_N,
                               date = WQ_monthly_complete$Date)


#O2 overall monthly ts
monthly.O.ts <- ts(WQ_monthly_complete$Interp_Dissolved_O2, start=c(2016,8),
                   frequency=12)
monthly.O.decomp <- stl(monthly.O.ts, s.window="periodic")
monthly.O.components <- as.data.frame(monthly.O.decomp$time.series[,1:3])
monthly.O.components <- mutate(monthly.O.components,
                               observed = WQ_monthly_complete$Interp_Dissolved_O2,
                               date = WQ_monthly_complete$Date)

#pH overall monthly ts
monthly.pH.ts <- ts(WQ_monthly_complete$Interp_pH, start=c(2016,8),
                   frequency=12)
monthly.pH.decomp <- stl(monthly.pH.ts, s.window="periodic")
monthly.pH.components <- as.data.frame(monthly.pH.decomp$time.series[,1:3])
monthly.pH.components <- mutate(monthly.pH.components,
                               observed = WQ_monthly_complete$Interp_pH,
                               date = WQ_monthly_complete$Date)



#P pre vs. post monthly trend analysis
P.pre.ts <- ts(pre_ordinance_monthly$Interp_Phosphorous, start = c(2016,8),
               frequency=12)
P.pre.decomp <- stl(P.pre.ts, s.window="periodic") 
P.pre.components <- as.data.frame(P.pre.decomp$time.series[,1:3])
P.pre.components <- mutate(P.pre.components,
                            observed = pre_ordinance_monthly$Interp_Phosphorous,
                            date = pre_ordinance_monthly$Date)
P.pre.smk <- smk.test(P.pre.ts) #no significant trend


P.post.ts <- ts(post_ordinance_monthly$Interp_Phosphorous, start = c(2021,4),
                frequency=12)
P.post.decomp <- stl(P.post.ts, s.window = "periodic")
P.post.components <- as.data.frame(P.post.decomp$time.series[,1:3])
P.post.components <- mutate(P.post.components,
                            observed = post_ordinance_monthly$Interp_Phosphorous,
                            date = post_ordinance_monthly$Date)
P.post.smk <- smk.test(P.post.ts) #significant trend! 

#P pre vs post plot
P.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Phosphorous)) +
  geom_line(data=P.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2018-10-01"), y=0.02,
            label=paste0("SMK p-value = ", round(P.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=P.post.components, aes(x=date, y=observed)) +
  geom_line(data=P.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=0.04,
            label=paste0("SMK p-value = ", round(P.post.smk$p.value, 3), "*"),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Total Phosphorous (mg/L)",
    x="")

#N pre vs. post trend analysis
N.pre.ts <- ts(pre_ordinance_monthly$Interp_Total_N, start = c(2016,8),
               frequency=12)
N.pre.decomp <- stl(N.pre.ts, s.window="periodic") 
N.pre.components <- as.data.frame(N.pre.decomp$time.series[,1:3])
N.pre.components <- mutate(N.pre.components,
                            observed = pre_ordinance_monthly$Interp_Total_N,
                            date = pre_ordinance_monthly$Date)
N.pre.smk <- smk.test(N.pre.ts) 


N.post.ts <- ts(post_ordinance_monthly$Interp_Total_N, start = c(2021,4),
                frequency=12)
N.post.decomp <- stl(N.post.ts, s.window = "periodic")
N.post.components <- as.data.frame(N.post.decomp$time.series[,1:3])
N.post.components <- mutate(N.post.components,
                            observed = post_ordinance_monthly$Interp_Total_N,
                            date = post_ordinance_monthly$Date)
N.post.smk <- smk.test(N.post.ts)

#N pre vs post plot
N.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Total_N)) +
  geom_line(data=N.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2018-10-01"), y=0.8,
            label=paste0("SMK p-value = ", round(N.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=N.post.components, aes(x=date, y=observed)) +
  geom_line(data=N.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-01-01"), y=0.8,
            label=paste0("SMK p-value = ", round(N.post.smk$p.value, 3)),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Total Nitrogen (mg/L)",
    x="")

#O2 pre vs. post trend analysis
O.pre.ts <- ts(pre_ordinance_monthly$Interp_Dissolved_O2, start = c(2016,8),
               frequency=12)
O.pre.decomp <- stl(O.pre.ts, s.window="periodic") 
O.pre.components <- as.data.frame(O.pre.decomp$time.series[,1:3])
O.pre.components <- mutate(O.pre.components,
                            observed = pre_ordinance_monthly$Interp_Dissolved_O2,
                            date = pre_ordinance_monthly$Date)
O.pre.smk <- smk.test(O.pre.ts) 


O.post.ts <- ts(post_ordinance_monthly$Interp_Dissolved_O2, start = c(2021,4),
                frequency=12)
O.post.decomp <- stl(O.post.ts, s.window = "periodic")
O.post.components <- as.data.frame(O.post.decomp$time.series[,1:3])
O.post.components <- mutate(O.post.components,
                            observed = post_ordinance_monthly$Interp_Dissolved_O2,
                            date = post_ordinance_monthly$Date)
O.post.smk <- smk.test(O.post.ts)

#O2 pre vs post plot
O.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_Dissolved_O2)) +
  geom_line(data=O.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2019-09-01"), y=45,
            label=paste0("SMK p-value = ",round(O.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=O.post.components, aes(x=date, y=observed)) +
  geom_line(data=O.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=45,
            label=paste0("SMK p-value = ", round(O.post.smk$p.value, 3)),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  labs(
    y="Dissolved Oxygen (mg/L)",
    x="")

#pH pre vs. post trend analysis
pH.pre.ts <- ts(pre_ordinance_monthly$Interp_pH, start = c(2016,8),
               frequency=12)
pH.pre.decomp <- stl(pH.pre.ts, s.window="periodic") 
pH.pre.components <- as.data.frame(pH.pre.decomp$time.series[,1:3])
pH.pre.components <- mutate(pH.pre.components,
                            observed = pre_ordinance_monthly$Interp_pH,
                            date = pre_ordinance_monthly$Date)
pH.pre.smk <- smk.test(pH.pre.ts) 


pH.post.ts <- ts(post_ordinance_monthly$Interp_pH, start = c(2021,4),
                frequency=12)
pH.post.decomp <- stl(pH.post.ts, s.window = "periodic")
pH.post.components <- as.data.frame(pH.post.decomp$time.series[,1:3])
pH.post.components <- mutate(pH.post.components,
                            observed = post_ordinance_monthly$Interp_pH,
                            date = post_ordinance_monthly$Date)
pH.post.smk <- smk.test(pH.post.ts)

#pH pre vs post plot
pH.prevspost.plot <- ggplot() +
  geom_line(data=WQ_monthly_complete, aes(x=Date, y=Interp_pH)) +
  geom_line(data=pH.pre.components, aes(x=date, y=trend), color="#005AB5") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2019-03-01"), y=8.75,
            label=paste0("SMK p-value = ",round(pH.pre.smk$p.value, 3)), 
            color="#005AB5") +
  geom_line(data=pH.post.components, aes(x=date, y=observed)) +
  geom_line(data=pH.post.components, aes(x=date, y=trend), color="#DC3220") +
  geom_text(data=WQ_monthly_complete,
            x=as.Date("2023-05-01"), y=9.5,
            label=paste0("SMK p-value = ", round(pH.post.smk$p.value, 4),"*"),
            color="#DC3220") +
  geom_vline(xintercept = as.Date("2021-04-01"), lty=2) +
  ylim(6.5,10) +
  labs(
    y="pH")

#combined pre vs post trend analysis plot
plot_grid(P.prevspost.plot, N.prevspost.plot, O.prevspost.plot, pH.prevspost.plot,
          nrow=4)
```

```{r SMK, echo=FALSE, message=FALSE, warning=FALSE}
#pre vs post SMK results summary table
pre.stat.summary <- c(P.pre.smk$statistic, N.pre.smk$statistic, 
                   O.pre.smk$statistic, pH.pre.smk$statistic)
pre.pvalue.summary <- c(P.pre.smk$p.value, N.pre.smk$p.value,
                        O.pre.smk$p.value, pH.pre.smk$p.value)
post.stat.summary <- c(P.post.smk$statistic, N.post.smk$statistic,
                       O.post.smk$statistic, pH.post.smk$statistic)
post.pvalue.summary <- c(P.post.smk$p.value, N.post.smk$p.value,
                        O.post.smk$p.value, pH.post.smk$p.value)

prevspost.SMK.summary <- data.frame(
  "Pre-Ordinance_Z.Stat" = pre.stat.summary,
  "Pre-Ordinance_p-value" = pre.pvalue.summary,
  "Post-Ordinance_Z.Stat" = post.stat.summary,
  "Post-Ordinance_p-value" = post.pvalue.summary,
  row.names = c("Total_Phosphorous", "Total_Nitrogen", "Dissolved_Oxygen", "pH"))

#smk summary table
kable(prevspost.SMK.summary, 
      digits=c(3,3,3,3),
      caption="Results Summary of SMK on Pre- and Post-Ordinance Monthly Averages")
```

### Comparison of Monthly and Yearly Trends

```{r yearlyTS, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Parameter Yearly Averages"}
#yearly TS with combined plots
par(mfrow = c(2,2))
yearly.P.ts <- ts(yearly_averages$Mean_Phosphorous, 
                  start = 2016,
                  frequency = 1)
plot(yearly.P.ts,
    xlab = "Year",
    ylab = "Phosphorous (mg/L)")

yearly.N.ts <- ts(yearly_averages$Mean_Total_N, 
                  start = 2016,
                  frequency = 1)
plot(yearly.N.ts,
     xlab = "Year",
     ylab = "Nitrogen (mg/L)")

yearly.O.ts <- ts(yearly_averages$Mean_Dissolved_O2, 
                  start = 2016,
                  frequency = 1)
plot(yearly.O.ts,
      xlab = "Year",
     ylab = "Dissolved Oxygen (mg/L)")

yearly.pH.ts <- ts(yearly_averages$Mean_pH, 
                  start = 2016,
                  frequency = 1)
plot(yearly.pH.ts,
     xlab = "Year",
     ylab = "pH")
par(mfrow = c(1,1))
```

```{r Pyearlyvsmonthly, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Comparing Yearly Averages with Interpolated Monthly Averages: Total Phosphorous"}
#comparison of monthly interpolated TS & yearly non-interpolated TS
#Phosphorous
P.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Phosphorous)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  xlim(as.Date("2016-08-01"),as.Date("2024-12-01")) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Phosphorous (mg/L)",
    title = "Yearly Mean Total Phosphorous")
  
plot_grid(P.prevspost.plot, P.yearly.plot,
          nrow=2,
          align = "v")
```

```{r Nyearlyvsmonthly, echo=FALSE, fig.cap="Comparing Yearly Averages with Interpolated Monthly Averages: Total Nitrogen", message=FALSE, warning=FALSE}
#Nitrogen
N.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Total_N)) +
  geom_line() +
  geom_point() +
  xlim(as.Date("2016-08-01"),as.Date("2024-12-01")) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Nitrogen (mg/L)",
    title = "Yearly Mean Total Nitrogen")


plot_grid(N.prevspost.plot, N.yearly.plot,
          nrow=2,
          align = "v")
```

```{r Oyearlyvsmonthly, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Comparing Yearly Averages with Interpolated Monthly Averages: Dissolved Oxygen"}
O.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_Dissolved_O2)) +
  geom_line() +
  geom_point() +
  xlim(as.Date("2016-08-01"),as.Date("2024-12-01")) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "Dissolved Oxygen (mg/L)",
    title = "Yearly Mean Total Dissolved Oxygen")


plot_grid(O.prevspost.plot, O.yearly.plot,
          nrow=2,
          align = "v")
```

```{r pHyearlyvsmonthly, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Comparing Yearly Averages with Interpolated Monthly Averages: pH"}
pH.yearly.plot <- ggplot() +
  geom_line(data=yearly_averages, aes(x=Date, y=Mean_pH)) +
  geom_line() +
  geom_point() +
  xlim(as.Date("2016-08-01"),as.Date("2024-12-01")) +
  geom_vline(xintercept = as.Date("2021-04-01"), lty = 2) +
  #scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Year",
    y = "pH",
    title = "Yearly Mean pH")


plot_grid(pH.prevspost.plot, pH.yearly.plot,
          nrow=2,
          align = "v")
```

## Question 2: Are there significant differences among the yearly averages for A) Total Phosphorous, B) Total Nitrogen, C) Dissolved Oxygen, and D) pH?

### ANOVA on Yearly Means

```{r anova, echo=FALSE, message=FALSE, warning=FALSE}
#ANOVA
yearly.P.anova <- aov(data=WQ_monthly_complete, Mean_Phosphorous ~ Year)
#TukeyHSD(yearly.P.anova)
P.HSD <- HSD.test(yearly.P.anova, "Year", group=TRUE)

yearly.N.anova <- aov(data=WQ_monthly_complete, Mean_Total_N ~ Year)
#TukeyHSD(yearly.N.anova)
N.HSD <- HSD.test(yearly.N.anova, "Year", group=TRUE)

yearly.O.anova <- aov(data=WQ_monthly_complete, Mean_Dissolved_O2 ~ Year)
#TukeyHSD(yearly.O.anova)
O.HSD <- HSD.test(yearly.O.anova, "Year", group=TRUE)

yearly.pH.anova <- aov(data=WQ_monthly_complete, Mean_pH ~ Year)
#TukeyHSD(yearly.pH.anova)
pH.HSD <- HSD.test(yearly.pH.anova, "Year", group=TRUE)

#anova results table
P.anova.results <- summary(yearly.P.anova)[[1]]
N.anova.results <- summary(yearly.N.anova)[[1]]
O.anova.results <- summary(yearly.O.anova)[[1]]
pH.anova.results <- summary(yearly.pH.anova)[[1]]

anova.results <- bind_rows(
  P.anova.results,
  N.anova.results,
  O.anova.results,
  pH.anova.results)

anova.results <- anova.results %>%
  mutate(Parameter = c(
    "Total_Phosphorous", "Total_Phosphorous", "Total_Nitrogen", "Total_Nitrogen",
    "Dissolved_O2", "Dissolved_O2", "pH", "pH")) %>%
  select(Parameter, Df, `Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)`)
 
kable(anova.results,
      row.names = T,
      digits = c(1,1,3,3,3,5),
      caption="Results Summary of ANOVA on Yearly Averages")
```

### Tukey's HSD 

```{r Tukey, echo=FALSE, message=FALSE, warning=FALSE}
#HSD results into table
P.groups <- as.data.frame(P.HSD$groups)
P.groups <- P.groups[order(rownames(P.groups)),]
N.groups <- as.data.frame(N.HSD$groups)
N.groups <- N.groups[order(rownames(N.groups)),]
O.groups <- as.data.frame(O.HSD$groups)
O.groups <- O.groups[order(rownames(O.groups)),]
#pH.groups <- as.data.frame(pH.HSD$groups)
#pH.groups <- pH.groups[order(rownames(pH.groups)),]

parameter.anova.groups <- data.frame(
  "Phosphorous" = P.groups,
  "Nitrogen" = N.groups,
  "Dissolved_O2"= O.groups)
  #"pH" = pH.groups)

kable(parameter.anova.groups,
      col.names=c("Year",
                  "Phosphorous_Mean","P_Group",
                  "Nitrogen_Mean","N_Group",
                  "Diss_O2_Mean","O2_Group"),
                  #"pH_Mean","pH_Group"),
            align=c("r","c","r","c","r","c","r","c"),
      digits=c(3,1,3,1,3,1,3),
      caption="Results Summary of Tukey's HSD on Yearly Averages")
```

---

\newpage

# Summary and Conclusions

## Total Phosphorous

## Total Nitrogen

## Dissolved Oxygen

## pH

## Conclusion

----

\newpage

# References
